A* with Perfect Potentials

### 2020-04-15 paper/ch_potentials.tex
Rewrote Lazy Topocore

\section{Avoiding Potential Function Evaluations}

We observe, that a potential evaluation is always paired with a queue push operation.
By avoiding push operations we can therefore also avoid potential evaluations.

In this section, we slightly modify how A* works internally.
Our modified version maintains the following entities:
(1) a tentative distance $D$ for every node,
(2) a queue of nodes $Q$ weighted by $D[v]+\pi(v)$.

Initially, $Q$ only contains $s$.
$D$ is $+\infty$ everywhere except for $D[s]=0$.

Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
Denote by $x$ a node removed with the queue.
Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
Instead, a limited, queue-less backtracking search at $x$ is started.
The idea is to process local tree-like areas of the graph without queue interaction.
We exploit that backtracking is fast on trees.

The backtracking search continues until either the tentative distance at a node is lower than the length of the path found by the search or a stop node is encountered.
Every node that was already pushed is a stop node.
Nodes with degree 4 or more is a stop node (TODO: @Tim out-degree ?).
Every degree 3 node that is reached via another degree 3 node from the backtracking search start is a stop node.
If the search was stopped because of the tentative distance, then it is aborted.
If the search was stopped because a stop node was encountered, the stop node is added to the queue.
This implies a potential evaluation.
If the node is already in the queue, a decrease key operation is executed.
One can view this modifed A* algorithm as a lazy version of \cite{topocore} where the set of virtually contracted nodes is computed at query time.

\subsection{Avoid Dead-Ends using Largest Biconnected Component}

A lot of nodes in road networks lead to dead-ends.
Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
We employ an idea from \cite{TOPOCORE} to avoid these nodes entirely.
This reduces the number of explored nodes and thus fewer potentials are evaluated.

In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
For nodes in the core, $a_v=v$.
We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
The basic CH-potential preprocessing step is only executed for $G_C$.

The query phase is divided into four steps.
In the first step, we use Dijkstra's algorithm to explore the component that contains $s$ until $a_s$ is reached.
If this search finds $t$, then $s$ and $t$ are part of the same component.
An optimal path was found and therefore the algorithm stops early.
Otherwise, in the second step starts.
It consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
The final path is the concatenation of all three path parts.

>>>>

\section{Avoiding Potential Function Evaluations at Degree Two Nodes}

We observe, that a potential evaluation is always paired with a queue push operation.
By avoiding push operations, we can therefore also avoid potential evaluations.
In this section, we slightly modify how A* works internally.
Our objective is process chains of low degree nodes sequentially without putting them into the queue.

We denote the tentative distance array used in the A* algorithm by $D$.
When removing a node $v_1$ from the A* queue, we iterate over all outgoing edges $(v_1,v_2)$ of $v_1$ and process the chain started by $(v_1, v_2)$.
If $D[v_2] < D[v_1]+w_q(v_1,v_2)$, our algorithm stops processing the chain and continues with the next outgoing edge of $v_1$.
Otherwise, $D[v_2]$ is set to $D[v_1]+w_q(v_1,v_2)$ and processing the chain continues.
If $v_2$ has only one outgoing edge that leads to a node $v_3$ different from $v_1$, we process $v_3$ analogously.
This is repeated until a $v_k$ is reached such that either $D[v_{k-1}]+w_q(v_{k-1},v_k)\ge D[v_k]$ or $v_k$ has more than one successor different from $v_{k-1}$.
Our algorithm continues removing nodes from the queue until a node $x$ is removed for which $D[x]\ge D[t]$.

Our optimization can be described as node contraction performed lazily at query time.
If $G$ is undirected, all nodes with a degree two in $G$ except $s$ are lazily contracted.
As we only employ contraction during the query, we do not loose flexibility.
This contrasts with most hierarchical approaches, that use contraction in the preprocessing step.

Not pushing degree two nodes into the queue gives a moderate to large speedup depending on how many degree two nodes are part of the graph.
We can achieve a slightly larger speedup by also not pushing every degree three node into the queue.

\section{Avoiding Potential Function Evaluations at Degree Three Nodes}

In the previous section, we described how a lazy degree two node contraction can be used to accelerate an A* search.
Denote by $I$ the set of nodes that can be pushed into the A* queue.
Using the $I$ from the previous section, $G\setminus I$ is a disjoint set of paths, if $G$ is undirected.
For performance reasons, it is important that $G\setminus I$ results in a disjoint collection of simple graphs.
Fortunately, we can allow slightly more complex graphs than just paths.
In this section, we use stars where the center node has degree three.
This allows us to further increase the size of $I$, which in turn reduces the number of A* queue operations and potential evaluations.

Let $v_k$ be the last node of a chain as in the previous section.
If $v_k$ has two successor nodes $u$ and $w$ that differ from $v_{k-1}$ and $v_k$ is not part of the queue, then we do not push $v_k$.
Instead, we follow the chains started by $(v_k, u)$ and $(v_k, w)$ as described in the previous section.
